---
title: "PP_Capstone_Proposal"
Author: "Stephen Patti"
Date:   "2018-09-06"
output:
  md_document:
    variant: markdown_github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Examining the causes of Power Plant Outages with respect to Frequency, nmber of outages, and Severity, duration of outages.

This study will examine the causes of US power plant outages based on the following data from the North American Electrical Reliability Corporation (NERC) Generating Availability Data System (GADS).
* CC_GT_2011_2017.csv - Combined Cycle Power Plants
* GT_Jet_Engine_2013_2016.csv - Simple Cycle Power Plants
* Fossil_steam_2013_2017.csv - Fossil Fuel Fired Boiler Steam Power Plants

My goal is determine which types of equipment failures are most responsible for unplanned/forced outages for each of these three types of plants.  To do this, I will have to do further examine the field Cause_Code, using the metadata associated with each four character code and chunking it into higher level or less granular categories.


With respect to outage type, "Type" / column 4, the following should be ignored as they are planned or maintenance outages and derates.
* MO - Maintenance Outage
* PE - Planned Extension
* PD - Planned Derate
* PO - Planned Outage
* RS - Reserve Shutdown

The first wrangling step is to clean the column names.
```{r c_cycle_edit, s_cycle_edit, echo=FALSE}
library(tidyverse)
# clean column names for combined and simple cycle power plant data sets
# fossil fuel steam already cleaned

#   create data frame for combined cycle
c_cycle_edit <- read.csv("CC_GT_2011_2017.csv")

#   replace . or series of . with _
names(c_cycle_edit) <- gsub(x = names(c_cycle_edit),
                            pattern = "(\\.)+",
                            replacement = "_")

#   replace trailing _ with ""
names(c_cycle_edit) <- gsub(x = names(c_cycle_edit),
                            pattern = "(_)+$",
                            replacement = "")

#   create working combined cycle file
write.csv(c_cycle_edit, file = "c_cycle_work.csv")

#   create data frame for simple cycle
s_cycle_edit <- read.csv("GT_Jet_Engine_2013_2016.csv")

#   replace . or series of . with _
names(s_cycle_edit) <- gsub(x = names(s_cycle_edit),
                            pattern = "(\\.)+",
                            replacement = "_")

#   replace trailing _ with ""
names(s_cycle_edit) <- gsub(x = names(s_cycle_edit),
                            pattern = "(_)+$",
                            replacement = "")

#   create working combined cycle file
write.csv(s_cycle_edit, file = "s_cycle_work.csv")
```

I wanted to create standard POSIX.ct date fields from the character fields in the data frames.


```{r ff_steam_work, c_cycle_work, s_cycle_work, echo=FALSE}
#create a combined field for start date and start time, start_dt,
#and end date / end time, end_dt.

ff_steam_work <- ff_steam_work %>%
  mutate(start_dt = paste(StartDate, StartTime, sep = " ", collapse = NULL)) %>%
  mutate(start_dt = mdy_hms(start_dt))

ff_steam_work <- ff_steam_work %>%
  mutate(end_dt = paste(End_Date, End_Time, sep = " ", collapse = NULL)) %>%
  mutate(end_dt = mdy_hms(end_dt))

c_cycle_work <- c_cycle_work %>%
  mutate(start_dt = paste(StartDate, StartTime, sep = " ")) %>%
  mutate(start_dt = mdy_hm(start_dt))

c_cycle_work <- c_cycle_work %>%
  mutate(end_dt = paste(End_Date, End_Time, sep = " ")) %>%
  mutate(end_dt = mdy_hm(end_dt))

s_cycle_work <- s_cycle_work %>%
  mutate(start_dt = paste(StartDate, StartTime, sep = " ")) %>%
  mutate(start_dt = mdy_hm(start_dt))

s_cycle_work <- s_cycle_work %>%
  mutate(end_dt = paste(End_Date, End_Time, sep = " ")) %>%
  mutate(end_dt = mdy_hm(end_dt))
```

I created new data frames with only the desired fields.

```{r ff_steam_worka, c_cycle_worka, s_cycle_worka, echo=FALSE}
#create a new data frame with specific fields in new order

ff_steam_worka <- ff_steam_work %>%
  select(Util_Unit, start_dt, Type, Cause_Code, end_dt,
         Time_To_Repair, TBF_Period_Hrs, IA_PH, 
         TBF_Service_Hrs, IA_SH, X_Derate)

c_cycle_worka <- c_cycle_work %>%
  select(Util_Unit, start_dt, Type, Cause_Code, end_dt,
         Time_To_Repair, TBF_Period_Hrs, IA_PH, 
         TBF_Service_Hrs, IA_SH, X_Derate)

s_cycle_worka <- s_cycle_work %>%
  select(Util_Unit, start_dt, Type, Cause_Code, end_dt,
         Time_To_Repair, TBF_Period_Hrs, IA_PH, 
         TBF_Service_Hrs, IA_SH, X_Derate)
```

Since I am focusing on unplanned outages, I used the following to filter each data set (fossil fuel steam shown as an example) to include only unplanned derates: D1, D2, D3, start up failures: SF, and unplanned full outages: U1, U2, U3.

```{r ff_steam_df, cc_df, ss_df, echo=FALSE}

library(tidyverse)
library(lubridate)

# Filter for only forced, not planned or reserve shutdowns, outages and derates

type_filter <- c("D1", "D2", "D3", "D4", "SF", "U1", "U2", "U3")
ff_steam_df <- ff_steam_df %>%
  filter(ff_steam_df$Type %in% type_filter)

cc_df <- cc_df %>%
  filter(cc_df$Type %in% type_filter)

ss_df <- ss_df %>%

#I also calculated the mean by Type using these data frames per below.  It’s not
#surprising that U1 has the highest mean as a U1 is an outage for which the
#generating unit must be shut down immediately 

# Calculate mean Time_To_Repair for each outage Type
ff_steam_df %>% group_by(Type) %>%
  summarise(Mean_TTR_Type_ff = mean(Time_To_Repair, na.rm = TRUE)) %>%
  arrange(desc(Mean_TTR_Type_ff))

cc_df %>% group_by(Type) %>%
  summarise(Mean_TTR_Type_cc = mean(Time_To_Repair, na.rm = TRUE)) %>%
  arrange(desc(Mean_TTR_Type_cc))
         
#I created box-plot graphs by type, in which the x axis was outage type, Type, and
#the y axis was time to repair, Time_To_Repair, for each data frame.  I also created
#stat summary graphs, x axis was year, y axis was type, and faceted by outage type.
#Both show distributions with long right tails and high range outliers.

# Create a Box Plot graph, x = Type, y = Time_To_Repair
ggplot(data = ff_steam_df, mapping = aes(x = Type, y = Time_To_Repair)) +
  geom_boxplot()

ggplot(data = cc_df, mapping = aes(x = Type, y = Time_To_Repair)) +
  geom_boxplot()

# create a stat summary plot, x = year(start_st), y = time to repair
ggplot(data = ff_steam_df) +
  stat_summary(
    mapping = aes(x = year(start_dt), y = Time_To_Repair),
    fun.ymin = min,
    fun.ymax = max,
    fun.y = mean
  ) +
  facet_wrap(~ Type, nrow = 2)

ggplot(data = cc_df) +
  stat_summary(
    mapping = aes(x = year(start_dt), y = Time_To_Repair),
    fun.ymin = min,
    fun.ymax = max,
    fun.y = mean
  ) +
  facet_wrap(~ Type, nrow = 2)

ff_box_Ty_TTR
ff_stat_sum

cc_box_Ty_TTR
cc_stat_sum

```
Given that there are hudreds of cause codes, Cause_Code, for each type of power plant, I grouped them by plant system, System, base on the NERC GADS appendix files.

```{r ff_steam_dfb, cc_dfb, ss_dfb, echo=FALSE}
library(tidyverse)
library(lubridate)

ff_steam_dfb <- ff_steam_df

#Convert Cause_Code to numeric
#Create a new variable, System, that categorizes Cause_Code according to NERC GADS
#Appendix
ff_steam_dfb <- ff_steam_dfb %>%
  mutate(Cause_Code = as.numeric(Cause_Code)) %>%
  mutate(System = 
            if_else(Cause_Code>=10 & Cause_Code<=1999, "Boiler",
            if_else((Cause_Code>=3110 & Cause_Code<=3549 | Cause_Code>=3800 & Cause_Code<=3999), "Balance_of_Plant",
            if_else(Cause_Code>=3600 & Cause_Code<=3730, "Electrical",
            if_else(Cause_Code>=4000 & Cause_Code<=4499, "Steam_Turbine",
            if_else(Cause_Code>=4500 & Cause_Code<=5298, "Generator",
            if_else(Cause_Code>=8000 & Cause_Code<=8845, "PCE",
            if_else((Cause_Code>=9000 & Cause_Code<=9340 | Cause_Code==0), "External",
            if_else(Cause_Code>=9504 & Cause_Code<=9720, "Regulatory",
            if_else(Cause_Code>=9900 & Cause_Code<=9960, "Personnel",
            if_else((Cause_Code==2 | Cause_Code>=9990 & Cause_Code<=9991), "Inactive", "Performance")))))))))))

cc_dfb <- cc_df
cc_dfb <- cc_dfb %>% 
  mutate(Cause_Code = as.numeric(Cause_Code)) %>% 
  mutate(System = 
           if_else((Cause_Code>=111 & Cause_Code<=1999 | Cause_Code>=6000 & Cause_Code<=6183), "HRSG",
           if_else((Cause_Code>=3110 & Cause_Code<=3549 | Cause_Code>=3800 & Cause_Code<=3899 | Cause_Code>=3950 & Cause_Code<=3999 | Cause_Code==6299 | Cause_Code==6399), "Balance_of_Plant",
           if_else(Cause_Code>=3600 & Cause_Code<=3730, "Electrical",
           if_else(Cause_Code>=4000 & Cause_Code<=4499, "Steam_Turbine",
           if_else(Cause_Code>=4500 & Cause_Code<=4750, "Generator",
           if_else(Cause_Code>=5000 & Cause_Code<=5299, "Gas_Turbine",
           if_else(Cause_Code>=5400 & Cause_Code<=5699, "Jet_Engine",
           if_else(Cause_Code==6200, "Miscellaneous",
           if_else(Cause_Code>=7800 & Cause_Code<=7960, "Expander_Turbine",
           if_else(Cause_Code>=8700 & Cause_Code<=8845, "PCE",
           if_else((Cause_Code>=9000 & Cause_Code<=9340 | Cause_Code==0), "External",
           if_else(Cause_Code>=9504 & Cause_Code<=9720, "Regulatory",
           if_else(Cause_Code>=9900 & Cause_Code<=9960, "Personnel",
           if_else((Cause_Code>=9990 & Cause_Code<=9991 | Cause_Code==2), "Inactive", "Performance")))))))))))))))

```
My next step is to create dummy numeric variables for each System type. This should allow me to begin using additional statistical methods.
I used the following code to assign numeric variables to per above.

```{r echo=FALSE}
# Create binary numeric variables for System categories

ff_steam_dfb <- ff_steam_dfb %>% 
  mutate(BlrNum = as.numeric((System == "Boiler")),
         BoPNum = as.numeric((System == "Balance_of_Plant")),
         ElecNum = as.numeric((System == "Electrical")),
         STNum = as.numeric((System == "Steam_Turbine")),
         GenNum = as.numeric((System == "Generator")),
         PCENum = as.numeric((System == "PCE")),
         RegNum = as.numeric((System == "Regulatory")),
         PersNum = as.numeric((System == "Personnel")),
         InactNum = as.numeric((System == "Inactive")),
         PerfNum = as.numeric((System == "Performance")))

cc_dfb <- cc_dfb %>% 
  mutate(HRSGNum = as.numeric((System == "HRSG")),
         BoPNum = as.numeric((System == "Balance_of_Plant")),
         ElecNum = as.numeric((System == "Electrical")),
         STNum = as.numeric((System == "Steam_Turbine")),
         GenNum = as.numeric((System == "Generator")),
         GTNum = as.numeric((System == "Gas_Turbine")),
         JENum = as.numeric((System == "Jet_Engine")),
         MiscNum = as.numeric((System == "Miscellaneous")),
         ETNum = as.numeric((System == "Expander_Turbine")),
         PCENum = as.numeric((System == "PCE")),
         RegNum = as.numeric((System == "Regulatory")),
         PersNum = as.numeric((System == "Personnel")),
         InactNum = as.numeric((System == "Inactive")),
         PerfNum = as.numeric((System == "Performance")))
```
I examinded the following distribution plots in order to set probable maximal loss (PML) at > 10 days, 240 hours, and <= 30 days, 720 hours, as well as maximum forseeable loss (MFL) at greater that 30 days.

```{r echo = FALSE}
ggplot(data = ff_steam_dfb, aes(x = Time_To_Repair)) +
         geom_density()
       
ggplot(data = ff_steam_dfb, aes(x = Time_To_Repair, color = Type)) +
  geom_density() +
  facet_grid(System ~ .)


ggplot(data = cc_dfb, aes(x = Time_To_Repair)) +           
  geom_density()

ggplot(data = cc_dfb, aes(x = Time_To_Repair, color = Type)) + 
  geom_density() +
  facet_grid(System ~ .)
```

```{r echo=FALSE}
# Create probable maximal loss (PML): 10 days < Time_To_Repair <= 30 days
# Create maximum forseeable loss (MFL): Time_To_Repair > 30 days

ff_steam_dfb <- ff_steam_dfb %>% 
  mutate(PML = 
           if_else(Time_To_Repair > 240 & Time_To_Repair <= 720, "1", "0")) %>% 
  mutate(MFL = 
           if_else(Time_To_Repair > 720, "1", "0"))

cc_dfb <- cc_dfb %>% 
  mutate(PML = 
           if_else(Time_To_Repair > 240 & Time_To_Repair <= 720, "1", "0")) %>% 
  mutate(MFL = 
           if_else(Time_To_Repair > 720, "1", "0"))
```

I used the following code to experiment with logistic regression for predicting (or showing an association wit) PML events by System.

```{r echo=FALSE}
# Experimenting with logistic regression to determine which systems determine,
# or are associated with PML and MFL.
ff_steam_dfb <- ff_steam_dfb %>% 
  mutate(PML = factor(PML, levels = c("0","1"))) %>% 
  mutate(MFL = factor(MFL, levels = c("0","1")))

PML_mod_ffs <- glm(PML ~ BlrNum + BoPNum + ElecNum + STNum + GenNum + 
                     PCENum + RegNum + PersNum + InactNum + PerfNum, 
                   data = ff_steam_dfb, family = "binomial")

# Remove NA Variables
PML_mod_ffs2 <- glm(PML ~ BlrNum + BoPNum + ElecNum + STNum + GenNum + 
                     PersNum + PerfNum, data = ff_steam_dfb, family = "binomial")

# Remove PersNum
PML_mod_ffs3 <- glm(PML ~ BlrNum + BoPNum + ElecNum + STNum + GenNum + 
                      PerfNum, data = ff_steam_dfb, family = "binomial")

# Remove Elecnum
PML_mod_ffs4 <- glm(PML ~ BlrNum + BoPNum + STNum + GenNum + 
                      PerfNum, data = ff_steam_dfb, family = "binomial")


cc_dfb <- cc_dfb %>% 
  mutate(PML = factor(PML, levels = c("0","1"))) %>% 
  mutate(MFL = factor(MFL, levels = c("0","1")))

PML_mod_cc <- glm(PML ~ HRSGNum + BoPNum + ElecNum + STNum + GenNum + GTNum +  
                     JENum + MiscNum + ETNum + PCENum + RegNum + PersNum + 
                    InactNum + PerfNum, data = cc_dfb, family = "binomial")

# Remove NA Variables
PML_mod_cc2 <- glm(PML ~ HRSGNum + BoPNum + ElecNum + STNum + GenNum + GTNum +  
                    JENum + PersNum + PerfNum, data = cc_dfb, family = "binomial")

# Remove HRSGNum
PML_mod_cc3 <- glm(PML ~ BoPNum + ElecNum + STNum + GenNum + GTNum +  
                     JENum + PersNum + PerfNum, data = cc_dfb, family = "binomial")

# Remove PersNum
PML_mod_cc4 <- glm(PML ~ BoPNum + ElecNum + STNum + GenNum + GTNum +  
                     JENum + PerfNum, data = cc_dfb, family = "binomial")
```
Experimenting with logistic regression for MFL
```{r echo=FALSE}
#MFL Fossil Fuel Steam Turbine
MFL_mod_ffs <- glm(MFL ~ BlrNum + BoPNum + ElecNum + STNum + GenNum + 
                     PCENum + RegNum + PersNum + InactNum + PerfNum, 
                   data = ff_steam_dfb, family = "binomial")

# Remove NA Variables: PCENum, REGNum, and InactNum
MFL_mod_ffs2 <- glm(MFL ~ BlrNum + BoPNum + ElecNum + STNum + GenNum + 
                     PersNum + PerfNum,data = ff_steam_dfb, family = "binomial")

# Remove STNum
MFL_mod_ffs3 <- glm(MFL ~ BlrNum + BoPNum + ElecNum + GenNum + 
                      PersNum + PerfNum,data = ff_steam_dfb, family = "binomial")

# Remove BOPNum, checking for multiple colinearity
MFL_mod_ffs4 <- glm(MFL ~ BlrNum + ElecNum + GenNum + 
                      PersNum + PerfNum,data = ff_steam_dfb, family = "binomial")


# MFL Combined Cycle (cc)
MFL_mod_cc <- glm(PML ~ HRSGNum + BoPNum + ElecNum + STNum + GenNum + GTNum +  
                    JENum + MiscNum + ETNum + PCENum + RegNum + PersNum + 
                    InactNum + PerfNum, data = cc_dfb, family = "binomial")

# Remove NA Variables: MiscNum, ETNum, PCENum, RegNum, InactNum
MFL_mod_cc2 <- glm(PML ~ HRSGNum + BoPNum + ElecNum + STNum + GenNum + GTNum +  
                    JENum + PersNum + PerfNum, data = cc_dfb, family = "binomial")

# Remove HRSGNum, highest Pr(>|z|)
MFL_mod_cc3 <- glm(PML ~ BoPNum + ElecNum + STNum + GenNum + GTNum +  
                     JENum + PersNum + PerfNum, data = cc_dfb, family = "binomial")

# Remove PersGNum, next highest Pr(>|z|)
MFL_mod_cc4 <- glm(PML ~ BoPNum + ElecNum + STNum + GenNum + GTNum +  
                     JENum + PerfNum, data = cc_dfb, family = "binomial")
```
After discussion with my course Mentor, Dhiraj, we refined the logistic regression approach.

```{r echo = FALSE}
# Refining Logistic Regression approach after discussion with course Mentor,
# Dhiraj Khanna

# Fossil Fuel Steam

ff_steam_dfb <- ff_steam_dfb[complete.cases(ff_steam_dfb),]

ff_steam_dfb <- ff_steam_dfb %>% 
  mutate(PML = factor(PML, levels = c("0","1"))) %>% 
  mutate(MFL = factor(MFL, levels = c("0","1")))

ff_steam_dfb$System <- as.factor(ff_steam_dfb$System)

fit_ffs <- glm(PML ~ System, data = ff_steam_dfb, family = "binomial")
summary(fit_ffs)
exp(coef(fit_ffs))

# Combined Cycle

cc_dfb <- cc_dfb[complete.cases(cc_dfb),]

cc_dfb <- cc_dfb %>% 
  mutate(PML = factor(PML, levels = c("0","1"))) %>% 
  mutate(MFL = factor(MFL, levels = c("0","1")))

cc_dfb$System <- as.factor(cc_dfb$System)

fit_cc <- glm(PML ~ System, data = cc_dfb, family = "binomial")
summary(fit_cc)
exp(coef(fit_cc))
```
While this doesn't have much predictive value with respect to failure rates and Time_To_Repair, from an equipment breakdown insurance perspective, it does show which systems will most likely result in a PML breakdown event.  I could use this in cooperation with our actuarial department to refine our rating for these risks.

Dhiraj and I also experimented with K-means clustering on this data.

```{r echo = FALSE}
# Fossil Fuel Steam

#First select the variables of importance
ClData_ffs <- ff_steam_dfb %>% select(System, Time_To_Repair, TBF_Period_Hrs)
ClData_ffs <- ClData_ffs[complete.cases(ClData_ffs),]
ffs_df <- scale(ClData_ffs[-1])

wssplot <- function(data, nc=15, seed=144){
  wss <- (nrow(data)-1)*sum(apply(data,2,var))
  for (i in 2:nc){
    set.seed(seed)
    wss[i] <- sum(kmeans(data, centers=i)$withinss)}
  plot(1:nc, wss, type="b", xlab="Number of Clusters",
       ylab="Within groups sum of squares")}

wssplot(ffs_df) #Indicates use of 7 Clusters

library(NbClust)
ncF <- NbClust(ffs_df, min.nc=2, max.nc=15, method="kmeans")
fit.kmF <- kmeans(ffs_df, 7, nstart=25)
fit.kmF$size
fit.kmF$centers
aggregate(ClData_ffs[-1], by=list(cluster=fit.kmF$cluster), mean)
ct.kmF <- table(ClData_ffs$System, fit.kmF$cluster)
print(ct.kmF)

library(flexclust)
randIndex(ct.kmF)

#Using PAM
library(cluster)
set.seed(666)
fit.pamF <- pam(ClData_ffs[-1], k=7, stand=TRUE)


# Combined Cycle

#First select the variables of importance
ClData_cc <- cc_dfb %>% select(System, Time_To_Repair, TBF_Period_Hrs)
ClData_cc <- ClData_cc[complete.cases(ClData_cc),]
cc_df <- scale(ClData_cc[-1])

wssplot <- function(data, nc=15, seed=144){
  wss <- (nrow(data)-1)*sum(apply(data,2,var))
  for (i in 2:nc){
    set.seed(seed)
    wss[i] <- sum(kmeans(data, centers=i)$withinss)}
  plot(1:nc, wss, type="b", xlab="Number of Clusters",
       ylab="Within groups sum of squares")}

wssplot(cc_df) #Indicates use of 3 Clusters

ncC <- NbClust(cc_df, min.nc=2, max.nc=15, method="kmeans")
fit.kmC <- kmeans(cc_df, 3, nstart=25)
fit.kmC$size
fit.kmC$centers
aggregate(ClData_cc[-1], by=list(cluster=fit.kmC$cluster), mean)
ct.kmC <- table(ClData_cc$System, fit.kmC$cluster)
print(ct.kmC)

randIndex(ct.kmC)

#Using PAM
set.seed(666)
fit.pamC <- pam(ClData_cc[-1], k=3, stand=TRUE)
```
Based on the indicies at close to 0, as well as the clustering showing heavy concentrations in one or two clusters, these models are not very useful.

This would have been more useful with the following types of data.
* Hours running prior to failure
* Number of starts and stops in the years prior to and between failure
* Plant maintenance expenditures
* Plant overhaul and maintenance schedules: number of daily, weekly, monthly maintenance items, number of major overhauls performed, etc.


