---
title: "PP_Capstone_Proposal"
Author: "Stephen Patti"
Date:   "2018-08-13"
output:
  md_document:
    variant: markdown_github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Examining the causes of Power Plant Outages with respect to Frequency, nmber of outages, and Severity, duration of outages.

This study will examine the causes of US power plant outages based on the following data from the North American Electrical Reliability Corporation (NERC) Generating Availability Data System (GADS).
* CC_GT_2011_2017.csv - Combined Cycle Power Plants
* GT_Jet_Engine_2013_2016.csv - Simple Cycle Power Plants
* Fossil_steam_2013_2017.csv - Fossil Fuel Fired Boiler Steam Power Plants

My goal is determine which types of equipment failures are most responsible for unplanned/forced outages for each of these three types of plants.  To do this, I will have to do further examine the field Cause_Code, using the metadata associated with each four character code and chunking it into higher level or less granular categories.


With respect to outage type, "Type" / column 4, the following should be ignored as they are planned or maintenance outages and derates.
* MO - Maintenance Outage
* PE - Planned Extension
* PD - Planned Derate
* PO - Planned Outage
* RS - Reserve Shutdown

The first wrangling step is to clean the column names.
```{r c_cycle_edit, s_cycle_edit, echo=FALSE}
library(tidyverse)
# clean column names for combined and simple cycle power plant data sets
# fossil fuel steam already cleaned

#   create data frame for combined cycle
c_cycle_edit <- read.csv("CC_GT_2011_2017.csv")

#   replace . or series of . with _
names(c_cycle_edit) <- gsub(x = names(c_cycle_edit),
                            pattern = "(\\.)+",
                            replacement = "_")

#   replace trailing _ with ""
names(c_cycle_edit) <- gsub(x = names(c_cycle_edit),
                            pattern = "(_)+$",
                            replacement = "")

#   create working combined cycle file
write.csv(c_cycle_edit, file = "c_cycle_work.csv")

#   create data frame for simple cycle
s_cycle_edit <- read.csv("GT_Jet_Engine_2013_2016.csv")

#   replace . or series of . with _
names(s_cycle_edit) <- gsub(x = names(s_cycle_edit),
                            pattern = "(\\.)+",
                            replacement = "_")

#   replace trailing _ with ""
names(s_cycle_edit) <- gsub(x = names(s_cycle_edit),
                            pattern = "(_)+$",
                            replacement = "")

#   create working combined cycle file
write.csv(s_cycle_edit, file = "s_cycle_work.csv")
```

I wanted to create standard POSIX.ct date fields from the character fields in the data frames.


```{r ff_steam_work, c_cycle_work, s_cycle_work, echo=FALSE}
#create a combined field for start date and start time, start_dt,
#and end date / end time, end_dt.

ff_steam_work <- ff_steam_work %>%
  mutate(start_dt = paste(StartDate, StartTime, sep = " ", collapse = NULL)) %>%
  mutate(start_dt = mdy_hms(start_dt))

ff_steam_work <- ff_steam_work %>%
  mutate(end_dt = paste(End_Date, End_Time, sep = " ", collapse = NULL)) %>%
  mutate(end_dt = mdy_hms(end_dt))

c_cycle_work <- c_cycle_work %>%
  mutate(start_dt = paste(StartDate, StartTime, sep = " ")) %>%
  mutate(start_dt = mdy_hm(start_dt))

c_cycle_work <- c_cycle_work %>%
  mutate(end_dt = paste(End_Date, End_Time, sep = " ")) %>%
  mutate(end_dt = mdy_hm(end_dt))

s_cycle_work <- s_cycle_work %>%
  mutate(start_dt = paste(StartDate, StartTime, sep = " ")) %>%
  mutate(start_dt = mdy_hm(start_dt))

s_cycle_work <- s_cycle_work %>%
  mutate(end_dt = paste(End_Date, End_Time, sep = " ")) %>%
  mutate(end_dt = mdy_hm(end_dt))
```

I created new data frames with only the desired fields.

```{r ff_steam_worka, c_cycle_worka, s_cycle_worka, echo=FALSE}
#create a new data frame with specific fields in new order

ff_steam_worka <- ff_steam_work %>%
  select(Util_Unit, start_dt, Type, Cause_Code, end_dt,
         Time_To_Repair, TBF_Period_Hrs, IA_PH, 
         TBF_Service_Hrs, IA_SH, X_Derate)

c_cycle_worka <- c_cycle_work %>%
  select(Util_Unit, start_dt, Type, Cause_Code, end_dt,
         Time_To_Repair, TBF_Period_Hrs, IA_PH, 
         TBF_Service_Hrs, IA_SH, X_Derate)

s_cycle_worka <- s_cycle_work %>%
  select(Util_Unit, start_dt, Type, Cause_Code, end_dt,
         Time_To_Repair, TBF_Period_Hrs, IA_PH, 
         TBF_Service_Hrs, IA_SH, X_Derate)
```

I created data frames converting the Cause_Codes to character fields in the event I wanted to do a join with the NERC GADS appendix codes.

```{r ff_steam_exp, c_cycle_exp, s_cycle_exp, echo=FALSE}

#convert Cause_Code iteger field to 4 character string/character field

ff_steam_exp <- ff_steam_worka

ff_steam_exp <- ff_steam_exp %>%
  mutate(Cause_Code = as.character(Cause_Code)) %>%
  mutate(Cause_Code = if_else(nchar(Cause_Code) == 2, paste0("00", Cause_Code),
                              if_else(nchar(Cause_Code) == 3, paste0("0", Cause_Code),
                                      if_else(nchar(Cause_Code) == 1, paste0("000", Cause_Code), Cause_Code))))

write.csv(ff_steam_exp, file = "ff_steam_expt.csv")

c_cycle_exp <- c_cycle_worka

c_cycle_exp <- c_cycle_exp %>%
  mutate(Cause_Code = as.character(Cause_Code)) %>%
  mutate(Cause_Code = if_else(nchar(Cause_Code) == 2, paste0("00", Cause_Code),
                              if_else(nchar(Cause_Code) == 3, paste0("0", Cause_Code),
                                      if_else(nchar(Cause_Code) == 1, paste0("000", Cause_Code), Cause_Code))))

write.csv(c_cycle_exp, file = "c_cycle_expt.csv")

s_cycle_expx <- s_cycle_worka

s_cycle_exp <- s_cycle_expx %>%
  mutate(Cause_Code = as.character(Cause_Code)) %>%
  mutate(Cause_Code = if_else(nchar(Cause_Code) == 2, paste0("00", Cause_Code),
                              if_else(nchar(Cause_Code) == 3, paste0("0", Cause_Code),
                                      if_else(nchar(Cause_Code) == 1, paste0("000", Cause_Code), Cause_Code))))
write.csv(s_cycle_exp, file = "s_cycle_expt.csv")
```

Since I am focusing on unplanned outages, I used the following to filter each data set (fossil fuel steam shown as an example) to include only unplanned derates: D1, D2, D3, start up failures: SF, and unplanned full outages: U1, U2, U3.

```{r ff_steam_df, cc_df, ss_df, echo=FALSE}

library(tidyverse)
library(lubridate)

# Filter for only forced, not planned or reserve shutdowns, outages and derates

type_filter <- c("D1", "D2", "D3", "D4", "SF", "U1", "U2", "U3")
ff_steam_df <- ff_steam_df %>%
  filter(ff_steam_df$Type %in% type_filter)

cc_df <- cc_df %>%
  filter(cc_df$Type %in% type_filter)

ss_df <- ss_df %>%

#I also calculated the mean by Type using these data frames per below.  It’s not
#surprising that U1 has the highest mean as a U1 is an outage for which the
#generating unit must be shut down immediately 

# Calculate mean Time_To_Repair for each outage Type
ff_steam_df %>% group_by(Type) %>%
  summarise(Mean_TTR_Type_ff = mean(Time_To_Repair, na.rm = TRUE)) %>%
  arrange(desc(Mean_TTR_Type_ff))

cc_df %>% group_by(Type) %>%
  summarise(Mean_TTR_Type_cc = mean(Time_To_Repair, na.rm = TRUE)) %>%
  arrange(desc(Mean_TTR_Type_cc))

ss_df %>% group_by(Type) %>%
  summarise(Mean_TTR_Type_ss = mean(Time_To_Repair, na.rm = TRUE)) %>%
  arrange(desc(Mean_TTR_Type_ss))

         
#I created box-plot graphs by type, in which the x axis was outage type, Type, and
#the y axis was time to repair, Time_To_Repair, for each data frame.  I also created
#stat summary graphs, x axis was year, y axis was type, and faceted by outage type.
#Both show distributions with long right tails and high range outliers.

ff_box_Ty_TTR
ff_stat_sum

cc_box_Ty_TTR
cc_stat_sum

ss_box_Ty_TTR
ss_stat_sum
```

Given that there are hudreds of cause codes, Cause_Code, for each type of power plant, I grouped them by plant system, System, base on the NERC GADS appendix files.

```{r ff_steam_dfb, cc_dfb, ss_dfb, echo=FALSE}
library(tidyverse)
library(lubridate)

ff_steam_dfb <- ff_steam_df

#Convert Cause_Code to numeric
#Create a new variable, System, that categorizes Cause_Code according to NERC GADS
#Appendix
ff_steam_dfb <- ff_steam_dfb %>%
  mutate(Cause_Code = as.numeric(Cause_Code)) %>%
  mutate(System = 
            if_else(Cause_Code>=10 & Cause_Code<=1999, "Boiler",
            if_else((Cause_Code>=3110 & Cause_Code<=3549 | Cause_Code>=3800 & Cause_Code<=3999), "Balance_of_Plant",
            if_else(Cause_Code>=3600 & Cause_Code<=3730, "Electrical",
            if_else(Cause_Code>=4000 & Cause_Code<=4499, "Steam_Turbine",
            if_else(Cause_Code>=4500 & Cause_Code<=5298, "Generator",
            if_else(Cause_Code>=8000 & Cause_Code<=8845, "PCE",
            if_else((Cause_Code>=9000 & Cause_Code<=9340 | Cause_Code==0), "External",
            if_else(Cause_Code>=9504 & Cause_Code<=9720, "Regulatory",
            if_else(Cause_Code>=9900 & Cause_Code<=9960, "Personnel",
            if_else((Cause_Code==2 | Cause_Code>=9990 & Cause_Code<=9991), "Inactive", "Performance")))))))))))

cc_dfb <- cc_df
cc_dfb <- cc_dfb %>% 
  mutate(Cause_Code = as.numeric(Cause_Code)) %>% 
  mutate(System = 
           if_else((Cause_Code>=111 & Cause_Code<=1999 | Cause_Code>=6000 & Cause_Code<=6183), "HRSG",
           if_else((Cause_Code>=3110 & Cause_Code<=3549 | Cause_Code>=3800 & Cause_Code<=3899 | Cause_Code>=3950 & Cause_Code<=3999 | Cause_Code==6299 | Cause_Code==6399), "Balance_of_Plant",
           if_else(Cause_Code>=3600 & Cause_Code<=3730, "Electrical",
           if_else(Cause_Code>=4000 & Cause_Code<=4499, "Steam_Turbine",
           if_else(Cause_Code>=4500 & Cause_Code<=4750, "Generator",
           if_else(Cause_Code>=5000 & Cause_Code<=5299, "Gas_Turbine",
           if_else(Cause_Code>=5400 & Cause_Code<=5699, "Jet_Engine",
           if_else(Cause_Code==6200, "Miscellaneous",
           if_else(Cause_Code>=7800 & Cause_Code<=7960, "Expander_Turbine",
           if_else(Cause_Code>=8700 & Cause_Code<=8845, "PCE",
           if_else((Cause_Code>=9000 & Cause_Code<=9340 | Cause_Code==0), "External",
           if_else(Cause_Code>=9504 & Cause_Code<=9720, "Regulatory",
           if_else(Cause_Code>=9900 & Cause_Code<=9960, "Personnel",
           if_else((Cause_Code>=9990 & Cause_Code<=9991 | Cause_Code==2), "Inactive", "Performance")))))))))))))))

ss_dfb <- ss_df
ss_dfb <- ss_dfb %>% 
  mutate(Cause_Code = as.numeric(Cause_Code)) %>% 
  mutate(System =
           if_else(Cause_Code>=3600 & Cause_Code<=3730, "Electrical",
           if_else(Cause_Code>=3810 & Cause_Code<=3999, "Balance_of_Plant",
           if_else(Cause_Code>=4500 & Cause_Code<=4899, "Generator",
           if_else(Cause_Code>=5000 & Cause_Code<=5299, "Gas_Turbine",
           if_else(Cause_Code>=7800 & Cause_Code<=7960, "Expander_Turbine",
           if_else(Cause_Code>=8656 & Cause_Code<=8845, "PCE",
           if_else((Cause_Code>=9000 & Cause_Code<=9340 | Cause_Code==0), "External",
           if_else(Cause_Code>=9504 & Cause_Code<=9720, "Regulatory",
           if_else(Cause_Code>=9900 & Cause_Code<=9960, "Personnel",
           if_else((Cause_Code>=9990 & Cause_Code<=9991 | Cause_Code==2), "Inactive",
           if_else(Cause_Code>=9997 & Cause_Code<=9999, "Performance", "Unknown"))))))))))))
```
My next step is to create dummy numeric variables for each System type. This should allow me to begin using additional statistical methods.


